---
title: "Getting started with POLARIS"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with POLARIS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

POLARIS takes two matrices measured on the same cells and returns three things
from one closed-form decomposition:

1. a **per-cell concordance score**, saying how far the two modalities agree
   about each individual cell;
2. a **joint cell embedding** that keeps structure supported by either
   modality, rather than only what they share;
3. a **cross-modal feature embedding** that scores candidate feature pairs,
   with a standard error attached to every pair.

Nothing here is fit separately. One call produces all three.

This vignette has two halves. The first runs on simulated data and executes when
the vignette is built, so every number you see is real. The second is a complete
walkthrough on a public multiome dataset; those blocks are shown but not run,
because they need a multi-gigabyte download and several minutes of compute.

```{r}
library(POLARIS)
```

# A worked example on simulated data

`polarisSimulate()` builds two modalities sharing a low-rank structure, where a
known subset of cells is driven by private rather than shared factors. Those
cells are cross-modality discordant by construction, which gives us a ground
truth to check the score against.

```{r}
set.seed(42)
sim <- polarisSimulate(n = 400, g = 80, p = 120, r = 5, discordant = 0.3)
dim(sim$X)
dim(sim$Y)
```

Both matrices are **cells in rows, features in columns**. This is the
orientation POLARIS expects, and it is the transpose of how Seurat stores
assays.

```{r}
fit <- Polaris(sim$X, sim$Y, gene.chr.ref = sim$anno,
               x.sds = 0, verbose = FALSE)
fit
```

## The concordance score

```{r}
fit <- SkymapSimScore(fit)
summary(fit$concordance)
```

`concordance` is the cosine similarity of a cell's paired coordinates, rescaled
to $[0,1]$ as $\frac{1}{2}(\cos + 1)$. Use this rather than
`simscore.cosine`, which is the raw cosine in $[-1,1]$; the two rank cells
identically but only the rescaled form is bounded.

The discordant cells should score lower, and they do:

```{r}
tapply(fit$concordance, ifelse(sim$shared < 0.5, "discordant", "concordant"),
       mean)
```

## Feature linkage

`SkymapLinkageTable()` enumerates candidate pairs within a genomic window of
each gene's transcription start site and scores them.

```{r}
tab <- SkymapLinkageTable(fit, gene.anno = sim$anno, window = 1e6,
                          verbose = FALSE)
head(tab[, c("gene", "peak", "dist", "score_raw", "SE", "score_adj")])
```

`SE` is a closed-form leave-one-cell-out jackknife. It does not refit the
model: because the score is bilinear in the fitted embedding, all *n*
omitted-cell scores follow from two matrix-vector products.

## Ranking by `z_dev`, not by magnitude

Raw scores decay with genomic distance, so ranking by `score_raw` recovers
proximity rather than regulation. `SkymapZdev()` standardizes within fixed
distance strata:

```{r}
tab <- SkymapZdev(tab)
head(tab[order(-tab$z_dev), c("gene", "peak", "dist", "distbin", "z_dev")])
```

This is the statistic the POLARIS benchmarks rank by. Distance-stratified
analyses also exclude pairs closer than 10 kb, which is not done for you:

```{r}
tab <- tab[tab$dist >= POLARIS_MIN_DIST, ]
nrow(tab)
```

`z_dev` is a variance-standardized **ranking** statistic. It is not a
calibrated probability or an FDR: the standard error is conditional on the
fitted model and computed on the denoised modalities.

# A complete run on real data

Everything below is the full pipeline on a public 10x PBMC multiome dataset,
from download to a ranked linkage table. The blocks are not executed when this
vignette is built, so copy them into a session to follow along.

The dataset is distributed through `SeuratData` as two objects measured on the
same 11,909 cells: RNA counts for 36,601 genes, and ATAC counts for 108,377
peaks. We keep the roughly 10,400 cells the original analysis annotated, which is
the whole experiment rather than a toy subset.

## Getting the data

```r
# SeuratData is not on CRAN, so install it from GitHub once. Its current
# release requires Seurat 5; POLARIS itself works with Seurat 4 or 5.
# install.packages("remotes")
# remotes::install_github("satijalab/seurat-data")

library(SeuratData)

InstallData("pbmcMultiome")                          # one-time download

pbmc.rna  <- LoadData("pbmcMultiome", "pbmc.rna")    # 36,601 genes  x 11,909 cells
pbmc.atac <- LoadData("pbmcMultiome", "pbmc.atac")   # 108,377 peaks x 11,909 cells
```

The peaks are already named `chr1-10109-10357`. POLARIS parses that form
directly, so no renaming is needed. `pbmc.rna` also carries
`seurat_annotations`, the cell type labels from the original analysis, which we
use at the end.

## Assembling one object

POLARIS needs both modalities on the same cells, in the same order. We build the
object from the raw count matrices so that nothing depends on which Seurat
version wrote the files.

```r
library(Seurat)
library(Signac)
library(POLARIS)

# The label "filtered" marks cells the original analysis rejected at QC. Dropping
# those, and any cell without a label, leaves 10,412 of the 11,909.
cells <- intersect(colnames(pbmc.rna), colnames(pbmc.atac))
ann   <- pbmc.rna$seurat_annotations[cells]
cells <- cells[!is.na(ann) & as.character(ann) != "filtered"]

# To try the workflow quickly first, subsample here and raise it later:
# set.seed(42); cells <- sample(cells, 3000)

rna.counts  <- SeuratObject::LayerData(pbmc.rna,  assay = "RNA",  layer = "counts")[, cells]
atac.counts <- SeuratObject::LayerData(pbmc.atac, assay = "ATAC", layer = "counts")[, cells]

obj <- CreateSeuratObject(counts = rna.counts, assay = "RNA")
# sep tells Signac how the peak names are punctuated: chr1-10109-10357
obj[["ATAC"]] <- Signac::CreateChromatinAssay(counts = atac.counts, sep = c("-", "-"))
obj$celltype  <- as.character(pbmc.rna$seurat_annotations[cells])
```

## Preprocessing, which POLARIS does not do for you

Normalization is a methodological choice, so POLARIS takes matrices that are
already normalized. These are the conventions used in the POLARIS paper.

```r
DefaultAssay(obj) <- "RNA"
obj <- NormalizeData(obj, normalization.method = "LogNormalize",
                     scale.factor = 1e6)

# ScaleData runs on ALL genes, not on the variable features. POLARIS selects its
# own feature space through the x.sds variance floor, which is deliberately
# broader than a highly-variable-gene list. Variable features are for an
# RNA-only PCA elsewhere; they are not used to build X.
obj <- ScaleData(obj, features = rownames(obj))

DefaultAssay(obj) <- "ATAC"
obj <- Signac::RunTFIDF(obj)
```

## Orientation, the one thing people get wrong

```r
# X and Y are CELLS IN ROWS. Seurat stores assays features-by-cells, so both
# need transposing. X is the scaled RNA; Y is the TF-IDF normalized ATAC.
X <- Matrix::t(SeuratObject::LayerData(obj, assay = "RNA",  layer = "scale.data"))
Y <- Matrix::t(SeuratObject::LayerData(obj, assay = "ATAC", layer = "data"))

dim(X)   # 10,412 cells x 36,601 genes, dense, about 2.8 GB
dim(Y)   # 10,412 cells x 108,377 peaks, sparse
```

Keeping `Y` sparse matters. POLARIS picks a different truncated-SVD solver for a
sparse second modality, and densifying 108,377 peaks would cost memory for
nothing.

## Gene annotation

The annotation must match the genome build the peaks were called against, and it
must carry strand, because transcription start sites are placed with a
strand-aware resize.

```r
# Bundled Cell Ranger ARC GRCh38 table, which matches this dataset.
anno <- polarisTSSRef("hg38")

# Alternatives: the annotation already attached to a Signac assay, which is the
# safest source because it is guaranteed to match your peaks, or one built from
# any TxDb, EnsDb or GTF.
# anno <- Signac::Annotation(obj[["ATAC"]])
# anno <- polarisMakeTSSRef(EnsDb.Hsapiens.v86::EnsDb.Hsapiens.v86)
# anno <- polarisMakeTSSRef("path/to/genes.gtf.gz")
```

## The fit

```r
# mc.cores parallelizes the per-chromosome feature decomposition only.
# gene.chr.ref is what makes linkage cis; leave it NULL for modality pairs with
# no genomic coordinates, such as CITE-seq.
fit <- Polaris(X, Y, gene.chr.ref = anno, mc.cores = 4, verbose = TRUE)
```

With `verbose = TRUE` this narrates what it decides, and those messages are
worth reading:

- how many features cleared each variance floor, which on this dataset keeps
  roughly a third of genes and nearly all peaks;
- the two modality ranks the denoiser chose, and the common minimum they are
  collapsed to. Both ranks come from the ratio rule, not from a fixed setting.
  Pass `reconcile.ranks = FALSE` to keep each modality's own rank;
- how many peaks fall outside the primary assembly, on `chrM`, scaffolds or
  patch contigs. Those are excluded from the per-chromosome step and reported
  rather than dropped silently.

The result is a `skymap`, a plain named list with a `print` method:

```r
fit                        # dimensions, selected ranks, what is stored

fit$skymap.cell            # cells x (r3 + r4), the joint embedding
fit$skymap.U               # cells x r3, the X-side paired coordinates
fit$skymap.V               # cells x r4, the Y-side paired coordinates
fit$skymap.P               # genes x r5, genome-wide gene loadings
fit$skymap.Q               # peaks x r5, genome-wide peak loadings
fit$skymap.feature.chr     # per-chromosome P, Q and singular values
fit$input.param            # every threshold and rank actually used

polarisFitSummary(fit)     # the same settings as a one-row data frame
polarisSpectrum(fit)       # the singular values, to redraw the denoiser plot
```

## Per-cell concordance

```r
fit <- SkymapSimScore(fit)

summary(fit$concordance)

# Concordance varies systematically across populations, so it is worth looking
# at by cell type rather than in aggregate.
sort(tapply(fit$concordance, obj$celltype[rownames(fit$skymap.cell)], median))
```

Two cautions. Concordance is not a quality score, and a low value is not a bad
cell: resting populations genuinely agree less across modalities than
transitioning ones. It is also sensitive to how abundant a population is in the
sample, so differences between a common and a rare cell type are not purely
biological. Compare like with like.

## Cis linkage with standard errors

```r
# Enumerate gene-peak pairs within 1 Mb of each TSS and score them. On the full
# 10,412 cells this gives about 1.8 million pairs.
tab <- SkymapLinkageTable(fit, gene.anno = anno, window = 1e6, mc.cores = 16)

colnames(tab)
# gene, peak, chr, dist, score_raw, SE, score_adj, p_lfsr, q_lfsr
```

`score_raw` is the weighted inner product of the gene and peak loadings. `SE` is
the leave-one-cell-out jackknife. `score_adj`, `p_lfsr` and `q_lfsr` come from
an empirical-Bayes shrinkage of the scores, useful for reporting but not for
ranking.

Pass `se = FALSE` to skip the standard errors, which is much cheaper if you only
want scores.

## Ranking

```r
tab <- SkymapZdev(tab)      # adds distbin, mu0, z_dev and z_dev_fl

# Always apply the distance floor before ranking. It is not done for you.
tab <- tab[tab$dist >= POLARIS_MIN_DIST, ]

head(tab[order(-tab$z_dev), c("gene", "peak", "dist", "score_raw", "SE", "z_dev")])
```

Compare that with `order(-tab$score_raw)`, and the two lists barely overlap.
`z_dev` differs from raw magnitude in two ways at once. It subtracts the mean
score of the pair's distance stratum, and those means fall steadily with
separation, so a pair competes only against others at a similar distance. It
then divides by the pair's own standard error, which demotes large but poorly
determined scores. In practice the second term does most of the reordering: the
highest raw scores tend to be strongly expressed loci whose scores are also the
least precisely estimated, and they drop sharply once divided by their error.

## Cell-type-specific linkage

Restricting the analysis to one population gives linkage specific to it, because
whole-population scores are an average in which cell-type signal is diluted. The
supported route is to fit that population:

```r
mono  <- colnames(obj)[obj$celltype == "CD14 Mono"]
fit.m <- Polaris(X[mono, ], Y[mono, ], gene.chr.ref = anno, mc.cores = 4)
tab.m <- SkymapZdev(SkymapLinkageTable(fit.m, gene.anno = anno, mc.cores = 4))
```

Note what does **not** work, because it fails quietly. Restricting
`fit$skymap.cell` to a subset of cells and reusing the global fit leaves
`score_raw` bit-for-bit unchanged, since the score is read from
`skymap.feature.chr`, which `Polaris()` built once on all cells. The standard
error does respond to the subset. On this dataset, restricting to the 721 CD14
monocytes moves the median `SE` from 5.32 to 1.32 while every score stays
identical, so the resulting `z_dev` divides a score computed on 3,000 cells by an
error computed on 721. Nothing errors; the ranking is simply wrong. Fit the
population instead, as above.

The POLARIS paper's cell-type panels use a third route, reusing the global cell
embedding and recomputing only the feature projection on the subset. That is not
exposed by this package.

## What to expect it to cost

Measured on all 10,412 cells with 16 cores: preprocessing about 20 seconds, the
fit about 8 minutes, the linkage table with standard errors about 4 minutes, and
a peak memory footprint near 21 GB.

Memory, not time, is the thing to plan around. It is dominated by the dense
scaled RNA matrix, 2.8 GB here, and by the standard-error step, which
reconstructs a cells-by-peaks block. Pass `se = FALSE` to skip that step when
memory is tight, or subsample cells as suggested above. A 3,000-cell subsample
runs in roughly two minutes with a much smaller footprint, which is a reasonable
way to rehearse the workflow before committing to the full dataset.

# Working from a Seurat object directly

`RunPolaris()` assembles the arguments and returns the joint embedding as a
`DimReduc`, so it drops into the usual Seurat workflow:

```r
res <- RunPolaris(obj, assay.x = "RNA", assay.y = "ATAC", mc.cores = 4)

res$object <- Seurat::FindNeighbors(res$object, reduction = "polaris",
                                    dims = seq_len(ncol(res$skymap$skymap.cell)))
res$object <- Seurat::FindClusters(res$object, resolution = 0.4)

res$skymap <- SkymapSimScore(res$skymap)
res$object$concordance <- res$skymap$concordance[colnames(res$object)]
```

It takes the layers POLARIS expects by default, `scale.data` for RNA and `data`
for ATAC, transposes them for you, and picks up the annotation attached to the
ATAC assay when there is one. `Polaris()` on plain matrices remains the
reproducible entry point; this only assembles its arguments.

# The `x.sds` floor

`Polaris()` keeps features of $X$ whose standard deviation exceeds `x.sds`,
default `0.95`, and features of $Y$ above `y.sds`, default `0`. The asymmetry is
intentional: $X$ is expected to be z-scored, so a floor near 1 selects genes
retaining appreciable variance, while $Y$ is on a different scale and only
constant features are dropped.

**This default assumes $X$ is z-scored.** If you pass a log-normalized layer,
almost every gene falls below the floor and the fit stops with an empty $X$.
Pass `x.sds = 0` in that case, as the CITE-seq analyses do.

# Beyond RNA and ATAC

POLARIS is a general two-modality method. The genomic layer is optional:
`gene.chr.ref` and the per-chromosome decomposition exist only to make linkage
*cis*, and without them you still get the concordance score, the joint
embedding, and the genome-wide loadings in `skymap.P` and `skymap.Q`.

For CITE-seq, $Y$ is the CLR-normalized ADT matrix. Protein features have no
genomic coordinate, so `z_dev` is undefined for protein-gene pairs; score them
from the loadings with an unweighted cosine instead.

```r
fit <- Polaris(t(obj[["RNA"]]$scale.data), t(obj[["ADT"]]$data), x.sds = 0)
P <- fit$skymap.P; Q <- fit$skymap.Q
cosine <- (P / sqrt(rowSums(P^2))) %*% t(Q / sqrt(rowSums(Q^2)))
```

# Notes on performance

`Polaris()` is CPU-only. Its dominant cost is the two truncated SVDs, which are
BLAS-bound for a dense $X$. The difference between R's reference BLAS and a
threaded one is large, on the order of 25x for that step, so linking R against
OpenBLAS or Accelerate is the single most effective speedup available.

`mc.cores` parallelizes the per-chromosome decomposition with `mclapply`, which
is fork-based and therefore serial on Windows.

Fits are deterministic. The truncated SVD used for $X$ draws a random starting
vector, so POLARIS fixes a seed internally and restores the previous RNG state
afterwards, leaving your own random stream untouched.

```{r}
sessionInfo()
```
