---
title: "Getting started with POLARIS"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with POLARIS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

POLARIS takes two matrices measured on the same cells and returns three things
from one closed-form decomposition:

1. a **per-cell concordance score**, saying how far the two modalities agree
   about each individual cell;
2. a **joint cell embedding** that keeps structure supported by either
   modality, rather than only what they share;
3. a **cross-modal feature embedding** that scores candidate feature pairs,
   with a standard error attached to every pair.

Nothing here is fit separately. One call produces all three.

```{r}
library(POLARIS)
```

# A worked example on simulated data

`polarisSimulate()` builds two modalities sharing a low-rank structure, where a
known subset of cells is driven by private rather than shared factors. Those
cells are cross-modality discordant by construction, which gives us a ground
truth to check the score against.

```{r}
set.seed(42)
sim <- polarisSimulate(n = 400, g = 80, p = 120, r = 5, discordant = 0.3)
dim(sim$X)
dim(sim$Y)
```

Both matrices are **cells in rows, features in columns**. This is the
orientation POLARIS expects, and it is the transpose of how Seurat stores
assays.

```{r}
fit <- Polaris(sim$X, sim$Y, gene.chr.ref = sim$anno,
               x.sds = 0, verbose = FALSE)
fit
```

## The concordance score

```{r}
fit <- SkymapSimScore(fit)
summary(fit$concordance)
```

`concordance` is the cosine similarity of a cell's paired coordinates, rescaled
to $[0,1]$ as $\frac{1}{2}(\cos + 1)$. Use this rather than
`simscore.cosine`, which is the raw cosine in $[-1,1]$; the two rank cells
identically but only the rescaled form is bounded.

The discordant cells should score lower, and they do:

```{r}
tapply(fit$concordance, ifelse(sim$shared < 0.5, "discordant", "concordant"),
       mean)
```

## Feature linkage

`SkymapLinkageTable()` enumerates candidate pairs within a genomic window of
each gene's transcription start site and scores them.

```{r}
tab <- SkymapLinkageTable(fit, gene.anno = sim$anno, window = 1e6,
                          verbose = FALSE)
head(tab[, c("gene", "peak", "dist", "score_raw", "SE", "score_adj")])
```

`SE` is a closed-form leave-one-cell-out jackknife. It does not refit the
model: because the score is bilinear in the fitted embedding, all *n*
omitted-cell scores follow from two matrix-vector products.

## Ranking by `z_dev`, not by magnitude

Raw scores decay with genomic distance, so ranking by `score_raw` recovers
proximity rather than regulation. `SkymapZdev()` standardizes within fixed
distance strata:

```{r}
tab <- SkymapZdev(tab)
head(tab[order(-tab$z_dev), c("gene", "peak", "dist", "distbin", "z_dev")])
```

This is the statistic the POLARIS benchmarks rank by. Distance-stratified
analyses also exclude pairs closer than 10 kb, which is not done for you:

```{r}
tab <- tab[tab$dist >= POLARIS_MIN_DIST, ]
nrow(tab)
```

`z_dev` is a variance-standardized **ranking** statistic. It is not a
calibrated probability or an FDR: the standard error is conditional on the
fitted model and computed on the denoised modalities.

# Working from real data

## Preprocessing, which POLARIS does not do for you

Normalization is a methodological choice, so POLARIS takes matrices that are
already normalized. The conventions used in the POLARIS paper are:

```r
obj <- NormalizeData(obj, assay = "RNA",
                     normalization.method = "LogNormalize",
                     scale.factor = 1e6)
obj <- FindVariableFeatures(obj, nfeatures = 3000)
obj <- ScaleData(obj, assay = "RNA", features = rownames(obj))
DefaultAssay(obj) <- "ATAC"
obj <- Signac::RunTFIDF(obj)
```

Note that `ScaleData` runs on **all** genes, not on the variable features.
POLARIS selects its own feature space through the `x.sds` variance floor, which
is deliberately broader than a highly-variable-gene list. The variable features
are used elsewhere (for an RNA-only PCA), not to build $X$.

## The `x.sds` floor

`Polaris()` keeps features of $X$ whose standard deviation exceeds `x.sds`,
default `0.95`, and features of $Y$ above `y.sds`, default `0`. The asymmetry is
intentional: $X$ is expected to be z-scored, so a floor near 1 selects genes
retaining appreciable variance, while $Y$ is on a different scale and only
constant features are dropped.

**This default assumes $X$ is z-scored.** If you pass a log-normalized layer,
almost every gene falls below the floor and the fit stops with an empty $X$.
Pass `x.sds = 0` in that case, as the CITE-seq analyses do.

## From a Seurat object

`RunPolaris()` assembles the arguments and returns the joint embedding as a
`DimReduc`, so it drops into the usual Seurat workflow:

```r
res <- RunPolaris(obj, assay.x = "RNA", assay.y = "ATAC", mc.cores = 4)

res$object <- Seurat::FindNeighbors(res$object, reduction = "polaris",
                                    dims = seq_len(ncol(res$skymap$skymap.cell)))
res$object <- Seurat::FindClusters(res$object, resolution = 0.4)

res$skymap <- SkymapSimScore(res$skymap)
res$object$concordance <- res$skymap$concordance[colnames(res$object)]
```

## Gene annotation

The annotation must match the genome build the peaks were called against. The
best source is the one already attached to your ATAC assay, which
`RunPolaris()` picks up automatically:

```r
anno <- Signac::Annotation(obj[["ATAC"]])
```

Otherwise use the bundled table, or build one for any other build:

```r
anno <- polarisTSSRef("hg38")
anno <- polarisMakeTSSRef(EnsDb.Hsapiens.v86::EnsDb.Hsapiens.v86)
anno <- polarisMakeTSSRef("path/to/genes.gtf.gz")
```

Strand matters. POLARIS places transcription start sites with a strand-aware
resize, so an unstranded annotation gives every minus-strand gene the wrong end
of the gene as its TSS. `polarisMakeTSSRef()` refuses input with no strand.

# Beyond RNA and ATAC

POLARIS is a general two-modality method. The genomic layer is optional:
`gene.chr.ref` and the per-chromosome decomposition exist only to make linkage
*cis*, and without them you still get the concordance score, the joint
embedding, and the genome-wide loadings in `skymap.P` and `skymap.Q`.

For CITE-seq, $Y$ is the CLR-normalized ADT matrix. Protein features have no
genomic coordinate, so `z_dev` is undefined for protein-gene pairs; score them
from the loadings with an unweighted cosine instead.

```r
fit <- Polaris(t(obj[["RNA"]]$scale.data), t(obj[["ADT"]]$data), x.sds = 0)
P <- fit$skymap.P; Q <- fit$skymap.Q
cosine <- (P / sqrt(rowSums(P^2))) %*% t(Q / sqrt(rowSums(Q^2)))
```

# Cell-type-specific linkage from one global fit

Because the joint embedding assigns a coordinate to every cell, feature
projections can be recomputed on any subset of cells **without refitting**.
Restricting $X$, $Y$ and $W$ to a population and passing them through the same
score gives population-specific linkage. Whole-population scores reflect an
average and carry little cell-type-specific signal; restricting to a cell type
is what recovers cell-type-specific links.

# Notes on performance

`Polaris()` is CPU-only. Its dominant cost is the two truncated SVDs, which are
BLAS-bound for a dense $X$. The difference between R's reference BLAS and a
threaded one is large — on the order of 25x for that step — so linking R
against OpenBLAS or Accelerate is the single most effective speedup available.

`mc.cores` parallelizes the per-chromosome decomposition with `mclapply`, which
is fork-based and therefore serial on Windows.

Fits are deterministic. The truncated SVD used for $X$ draws a random starting
vector, so POLARIS fixes a seed internally and restores the previous RNG state
afterwards, leaving your own random stream untouched.

```{r}
sessionInfo()
```
